---
title: "Analysis of ground reaction forces for the detection of musculoskeletal disorders"
author: "Carlos Felipe Rengifo Rodas"
date: '2022-06-01'
output: bookdown::pdf_document2
---

# Introduction

## Dataset and variables

GaitRec is a free to use database that comprises the ground reaction forces 
(GRF) of $2,084$ participants with musculoskeletal disorders at hip ($450$),
knee ($625$), ankle ($627$), and calcaneous ($382$), as well as $211$ healthy
controls. The GRF were recorded in the vertical, vertical, anteroposterior, and 
mediolateral directions, while the participants walked at self-select speed 
along a 10 $m$ long straight corridor. The GRF were measured using one force
plate for each foot (Kistler, Type 9281B12, Winterthur, CH).

GaitRec comprises raw and processed GRF. The processed forces were obtained as
follows:  (1) passing the raw GRF through a second-order low-pass Butterworth 
filter with a cut-off frequency of 20 Hz to reduce noise, (2) normalizing the 
duration of each stride to $100\%$, and (3) subsampling the GRF to obtain $101$ 
points per stride. Unlike the raw GRF, which were stored in Newton, the 
processed GRF were expressed as a multiple of the body weight to avoid bias when 
comparing the GRF of participants of different mass. Each processed GRF file is 
a $104$-column matrix. The first three columns are the participant, session, and 
stride identifiers, respectively, and the resting $101$ columns are the 
normalized GRF.

Besides the GRF, GaitRec also comprises a metadata file with subject's, 
identifier, sex, age, height, weight, affected side (right, left, or none), and
the class to which the participant belongs, which can be healthy control (HC), 
hip (H), knee (K), ankle (A), or calcaneus (C).

## Goal of the project

The objective of the present project is to compare different combinations of 
predictors and multiclass machine learning models in terms of their accuracy, 
sensitivity, and specificity to classify GRF as belonging to a healthy control 
or to a participant with hip, knee, ankle, or calcaneal injury.

## Key steps

The main steps followed in this study were:

1. Bind the GRF to the class to which the participant belongs, resulting in a 
single data frame containing the predictors (GRF) and the output variable.

2. Divide the data into training and test sets according to the partition 
proposed by the GaitRed authors. This partitioning is important because it 
ensures comparability between different studies.

3. Extract features from the dataset. In this study, we considered three types 
of predictors: (1) mean and standard deviation of the GRF in each direction 
(vertical, anteroposterior, and mediolateral), (2) temporal location and 
amplitude of the GRF's maxima and minima, and (3) the principal component of the 
GRF.

# Methods

## Data loading and preprocessing

The following code checks if the libraries required to run the analysis of the 
GRF are installed.

```{r, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
# Tidyverse package
if(!require(tidyverse)) 
  install.packages("tidyverse", repos="http://cran.us.r-project.org")

# Caret package
if(!require(caret)) 
  install.packages("caret", repos="http://cran.us.r-project.org")

# Ggplot2 package
if(!require(ggplot2)) 
  install.packages("ggplot2", repos="http://cran.us.r-project.org")

# Kernlab package
if(!require(kernlab)) 
  install.packages("kernlab", repos="http://cran.us.r-project.org")

# Knitr package
if(!require(knitr)) 
  install.packages("knitr", repos="http://cran.us.r-project.org")

# Readxl package
if(!require(readxl)) 
  install.packages("readxl", repos="http://cran.us.r-project.org")
options(digits = 3)
```

The following code checks whether the files of the GaitRec database are stored 
in the *dataset* folder. If these files are not found, they are downloaded from
the GaitRec website.

```{r, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
# Read the CSV files with ground reaction forces in the vertical, anteroposterior,
# and mediolateral directions:

# Vertical ground reaction forces
if (file.exists("dataset/GRF_F_V_PRO_left.csv")) {
  grf_v_left <- read.csv(file = "dataset/GRF_F_V_PRO_left.csv")   
} else {
  download.file("https://springernature.figshare.com/ndownloader/files/22063191",
                "dataset/GRF_F_V_PRO_left.csv")
}

# Anterior-posterior ground reaction forces
if (file.exists("dataset/GRF_F_AP_PRO_left.csv")) {
  grf_ap_left <- read.csv(file = "dataset/GRF_F_AP_PRO_left.csv")   
} else {
  download.file("https://springernature.figshare.com/ndownloader/files/22063185",
                "dataset/GRF_F_AP_PRO_left.csv")
}

# Medio-lateral ground reaction forces
if (file.exists("dataset/GRF_F_ML_PRO_left.csv")) {
  grf_ml_left <- read.csv(file = "dataset/GRF_F_ML_PRO_left.csv")   
} else {
  download.file("https://springernature.figshare.com/ndownloader/files/22063113",
                "dataset/GRF_F_ML_PRO_left.csv")
}
```

In the sake of clarity, the values $0$, $1$, $2$ and NaN that takes the 
AFFECTED_SIDE variable of the metadata file are replaced by the values LEFT, 
RIGHT, BOTH or NONE, respectively. 

```{r, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
meta <- read.csv(file = "dataset/GRF_metadata.csv") %>% 
          mutate(CLASS_LABEL = factor(CLASS_LABEL)) %>%
          mutate(AFFECTED_SIDE = case_when(AFFECTED_SIDE == 0 ~ "LEFT",
                                           AFFECTED_SIDE == 1 ~ "RIGHT",
                                           AFFECTED_SIDE == 2 ~ "BOTH",
                                           is.na(AFFECTED_SIDE) ~ "NONE")) %>%
          mutate(AFFECTED_SIDE = factor(AFFECTED_SIDE))
```

The GRF in the vertical, anteroposterior, and mediolateral directions are binded
with subject, session, and trial identifiers. Hence, *grf_left* is a 306-column 
matrix that contains the 303 GRF and three identifiers.

```{r, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
join_vars <- c("SUBJECT_ID","SESSION_ID","TRIAL_ID")
grf_left <- grf_v_left %>% inner_join(grf_ap_left,by = join_vars) %>%
                inner_join(grf_ml_left,by = join_vars) 

# Remove unused variables
#rm(grf_v_left,grf_ap_left,grf_ml_left)
```

*grf_left* is binded with the metadata to obtain the data frame *dataset*, which
will be used to apply the machine learning algorithms.

```{r, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
# Join the data frames "grf_left" and "meta", and then remove the users whose 
# affected side is the right or both
dataset <- meta %>% select(SUBJECT_ID, SESSION_ID, CLASS_LABEL, AFFECTED_SIDE, 
                           SESSION_TYPE, TRAIN_BALANCED, TEST) %>%
                    inner_join(grf_left, by = c("SUBJECT_ID","SESSION_ID")) %>% 
                    filter(AFFECTED_SIDE %in% c("LEFT","NONE")) %>%
                    filter(SESSION_TYPE == 1)

# Remove unused variables
#rm(grf_left)  
```

## Data exploration

The Table \@ref(tab:ppgroup) shows the number of participants per group.

```{r ppgroup, echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
meta %>% group_by(CLASS_LABEL) %>% summarise(n = n_distinct(SUBJECT_ID)) %>%
  rename("Class" = CLASS_LABEL, "Number of participants" = n) %>%
  knitr::kable(align = 'c')
```

The following code creates a function that calculates the mean and the 
confidence interval for the forces in a given direction (vertical, 
anteroposterior, or mediolateral)

```{r, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
ci_forces <- function(data,colforces) {
  # Rename the columns with the forces with the numbers from 1 to 101
  colnames(data)[colforces] <- 1:101
  data %>%
  gather(key = "TIME", value = "FORCE", all_of(colforces)) %>%
    mutate(TIME = as.numeric(TIME)) %>%
    group_by(CLASS_LABEL,TIME) %>%
    summarise(MEAN = mean(FORCE), 
              LOWER = MEAN - 1.96*sd(FORCE),
              UPPER = MEAN + 1.96*sd(FORCE)) %>%
    gather(key = "TYPE", value = "VALUE", c("MEAN","LOWER", "UPPER"))
}
```

The next step is to apply the *ci_forces* function to the GRF in each direction,
and then plot these forces per group.

```{r grfall, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE, out.width="75%", fig.align='center', fig.cap ="Lower, mean, and upper GRF per group. Healthy controls (HC), Hip (H), Knee (K), Ankle (A), and Calcaneous (C)."}
# Columns with the first and last forces in each direction
noforces  <- 1:8
f_v_1     <- which(colnames(dataset) == "F_V_PRO_1")
f_v_101   <- which(colnames(dataset) == "F_V_PRO_101")
f_ap_1    <- which(colnames(dataset) == "F_AP_PRO_1")
f_ap_101  <- which(colnames(dataset) == "F_AP_PRO_101")
f_ml_1    <- which(colnames(dataset) == "F_ML_PRO_1")
f_ml_101  <- which(colnames(dataset) == "F_ML_PRO_101")

# Confidence interval by direction
ci_forces_v  <- ci_forces(dataset,f_v_1:f_v_101) %>% mutate(DIRECTION = "V")
ci_forces_ap <- ci_forces(dataset,f_ap_1:f_ap_101) %>% mutate(DIRECTION = "AP")
ci_forces_ml <- ci_forces(dataset,f_ml_1:f_ml_101) %>% mutate(DIRECTION = "ML")
ci_forces_3D <- rbind(ci_forces_v, ci_forces_ap, ci_forces_ml) 

# Plot confidence intervals
ci_forces_3D %>% ggplot(aes(x = TIME,y = VALUE, color = CLASS_LABEL)) + 
  geom_line() + facet_grid(DIRECTION ~ TYPE, scales = "free_y")
```
According to Figure \@ref(fig:grfall), there are notable differences between the
healthy controls (HC) and the other four groups. On the other hand, 
Figure \@ref(fig:grfhk) shows that the GRF of the participants with hip (H) and 
knee (K) injuries are very similar in the anteroposterior and the vertical 
directions.

```{r grfhk, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE, out.width="75%", fig.align='center', fig.cap ="Lower, mean, and upper GRF for the participants with hip (H) and knee (K) injuries."}
# Plot confidence intervals
ci_forces_3D %>% filter(CLASS_LABEL %in% c("H","K")) %>%
  ggplot(aes(x = TIME,y = VALUE, color = CLASS_LABEL)) + 
  geom_line() + facet_grid(DIRECTION ~ TYPE, scales = "free_y")
```

The Figure \@ref(fig:grfac) shows that the GRF of the participants with 
ankle (A) and calcaneous (K) impairment are very similar in the mediolateral and 
the vertical directions.

```{r grfac, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE, out.width="75%", fig.align='center', fig.cap ="Lower, mean, and upper GRF for the participants with ankle (A) and calcaneous (C) injuries."}
# Plot confidence intervals
ci_forces_3D %>% filter(CLASS_LABEL %in% c("A","C")) %>%
  ggplot(aes(x = TIME,y = VALUE, color = CLASS_LABEL)) + 
  geom_line() + facet_grid(DIRECTION ~ TYPE, scales = "free_y")
```

## Training and test sets

The metadata file includes a column that defines whether a subject's GRF belongs 
to the training set or the test set. It is important to maintain this partition
to ensure comparability between different studies. The following chunk of code
creates the 303-columns matrices *xtest* and *xtrain* with the GRF in vertical, 
anteroposterior, and mediolateral directions.

```{r, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
# Training set
trainset <- dataset %>% filter(TRAIN_BALANCED == 1)
# Predictors for the training set
xtrain <- trainset %>% select(-SUBJECT_ID, -CLASS_LABEL, -AFFECTED_SIDE,
                         -SESSION_ID, -TRIAL_ID, -SESSION_TYPE, 
                         -TRAIN_BALANCED, -TEST) %>% as.matrix()
# Class for the training set                   
ytrain <- trainset$CLASS_LABEL
  
## Test set
testset <- dataset %>% filter(TEST == 1)
# Predictors for the test set
xtest <- testset %>% select(-SUBJECT_ID, -CLASS_LABEL, -AFFECTED_SIDE,
                            -SESSION_ID, -TRIAL_ID, -SESSION_TYPE, 
                            -TRAIN_BALANCED, -TEST) %>% as.matrix()
# Class for the test set                   
ytest <- testset$CLASS_LABEL
```

# Results

## Principal component analysis (PCA)

The following code calculates of principal components of GRF used for training.

```{r, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
# Principal component analysis
xtrain_pca <- prcomp(xtrain)
# Cumulative standard deviation
cumsdev <- cumsum(xtrain_pca$sdev/sum(xtrain_pca$sdev))
# Minimum number of columns required to explain 95% of the variability
max_pca_col <- min(which(cumsdev >= 0.95))
ztrain <- xtrain_pca$x[,1:max_pca_col]
```

The above calculations shows that the first `r max_pca_col` columns of the PCA 
matrix explain 95% of the variability of the training set. The Figure \@ref(fig:grfac) 
shows variability versus the number of columns.

```{r variability, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE, out.width="75%", fig.align='center', fig.cap ="Percentage of variability versus the number of columns of the matrix that results from applying PCA to the GRF."}
# Plot the cumulative standard deviation
data.frame(n = 1:303, cumsdev) %>% ggplot(aes(n,cumsdev)) + 
  scale_y_continuous(trans='log2') + geom_point() 
```
Principal component analysis for the test set. 

```{r, echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
# Principal component analysis
pca_test <- prcomp(xtest)
# Cumulative standard deviation
cumsdev <- cumsum(pca_test$sdev/sum(pca_test$sdev))
# Reduced set of features
ztest <- pca_test$x[,1:max_pca_col]
```

The confusion matrix for K nearest neighbors model shows that no matter the 
musculoskeletal disorder, a participant is likely to be classified as a healthy 
control (fourth row of the confusion matrix). This means that the classes (A, C, 
H, and K) have high specificity but very low sensitivity.  

```{r, echo=TRUE, error=FALSE, warning=FALSE, message=FALSE}
## K-NEAREST NEIGHBORS
model_knn <- train(ztrain, ytrain, method = "knn")
yhat_knn <- predict(model_knn, ztest)
cm_knn <- confusionMatrix(yhat_knn, ytest)
acc_knn <- cm_knn$overall["Accuracy"]
cm_knn$table %>% knitr::kable(align = 'c')
```


<!-- 
Considering the similarities between groups H and K (\@ref(fig:grfhk)) and
between the groups A and C (\@ref(fig:grfac)), we will combine the classes H and
K into a single class named hip-knee (HK), and the classes A and C into the 
class angle-calcaneous (AC).
-->